---
title: "Urban Big Data Analytics - Lecture 7 Group Session: Exploratory Data Analysis"
author: "Andy Hong"
date: "July 25, 2019"
output:
  html_document:
    css: my_styles.css
  pdf_document: default
  word_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

<br>

## Prerequisites

Please locate your `vsp_bigdata` folder under "My Documents" and navigate to `group-session`. Create `07-lecture` folder under the `group-session` folder.

1. For this group session, we will use the Seattle crime database. <br>
If you don't have the Seattle crime database, please download this CSV file and load it into Exploratory: <a href="http://urbanbigdata.github.io/05-lecture-spatial_data/group-session/seattle_crime_2013.csv" download> Seattle crime data 2013 </a><br>

2. Now, you need to download the US census data to add demographic information: <a href="http://urbanbigdata.github.io/05-lecture-spatial_data/group-session/seattle_census_2013.csv" download> Seattle Census ACS data 2013 </a><br>


## Instruction

### 1. Synopsis
The purpose of this group session is to get you familiar with **Exploratory Data Analysis (EDA)**. There are three steps in this group session. First, we will create a histogram for median household income. We will also create a bar plot using the crime types. <br>

Second, we will create a scatter plot using population and total income variables. We will also create a box plot using the crime type and the total number of crimes. We will also do a log-transformation of the total crime data.<br> 

Third, we will create a correlation coefficient plot using some of the key variables in the dataset.

<br>

### 2. Initial preparation

First, open Exploratory and import the Seattle crime data only for the year 2013. <br>

Now create a new column named `tract` using `mutate` and the the following calculation:
```{r, eval=F}

as.numeric(gsub("\\..*", "", CensusTract2000))

```

<br>
This calculation extracts only the tract number from the data. <br>
**Further information for advanced students**: the `gsub` function uses regular expression to filter and replace characters. In this case the regular expression finds all the characters that go after . and replace them with a blank character. For example, 6200.1043 becomes 6200.<br> You can read more about this function here: http://stat.ethz.ch/R-manual/R-devel/library/base/html/grep.html

Now, we will join the crime data with the census data. To do this, let's first group the data by `census tract`. Click `+` and choose `Group_by` and choose `tract`. And click `+` again and choose `Group_by` and choose  `SummarizedOffenseDescription`. `Group_by` is not going to do anything unless we summarize the data. Let's summarise the data and get the total number of crime by census tract. We can do this by clicking `+` and choose `Summarize` and choose `# Number of rows` for the Column, and give the New column name as `total_crime`.

Next, we will join the Seattle census data to add demographic information to the crime data. Add File Data by using the `+` sign and select `seattle_census_2013.csv` file. <br>

Now, go back to the Seattle crime data, and click `+` and select `Join`.<br>
In the target data frame, choose `seattle_census_2013` and for both the Current Column and Target Column, choose `tract`. We can see that Census data are joined to the Seattle crime data. <br><br><br>


### 3. First part - exploring one variable

Now, we are going to use EDA to explore one variable. Go to `Chart` and choose `Histogram` for the type. For X Axis, choose `med_hh_inc`. You will see a distribution of median household income in Seattle. <br>

Now, let's create another `Chart` by clicking `+`. This time, choose `Bar` for the type, and for X Axis, choose `SummarizedOffenseDescription` and for Y Axis, choose `total_crime`. You can usually leave the Y Axis as `Number of Rows`, but in this case we need to use the total crime. Once we are done, we can make the plot pretty by sorting the data by Y Axis and change the orientation from Vertical to Horizontal. 

<br><br>


### 4. Second part - Exploring two variables

Now, let's explore two variables. For this to happen, we will need to turn off the crime type. So, go back to the fourth step on your right panel, and remove `SummarizedOffenseDescription` and click OK. We will only group the data by `tract`. <br>

For the scatter plot, go to `Chart` and choose `Scatter` for the type. On X Axis, choose `pop` and for Y Axis, choose `total_crime`. We can see the relationship between population density and total crime. <br>

We can make the plot fancier by adding a trend line. Click the accordian-style menu next the Y Axis, and click `Trend Line` and choose `Linear Regression` for the type. It shows a beautiful scatter plot with a trend line showing the relationship between population and crime.<br>

Note that the distribution of crime data is skewed to the right, meaning there are values close to 0s. We can transform the crime data to overcome this non-normality issue. <br>

To do that, click `+` and choose `Mutate`. On the new Column name, type in `total_crime_log` and in the calculation field, type in `log(total_crime)`. Hit Run and we see that a new variable called `total_crime_log` was created. <br>

For the scatter plot, let's use `total_crime_log` for the Y Axis and see how the plot changes. <br>

Now, we will create anothe `Chart` by clicking `+`. For this to work, we will need to include crime type again for the grouping variable. Go back to the fourth step on the right panel, and click `Add New` below `tract` and choose  `SummarizedOffenseDescription`. <br>

Select the `Summarize` step on your right panel, and choose `BoxPlot` for the type. For the X Axis, choose `SummarizedOffenseDescription` and choose `total_crime` for the Y Axis. We can see that the boxplot is created. One problem is that there are too many crime types. <br><br>

Let's limit the crime types to top five crimes. We can do this by clicking `+` and choose `Filter` and select `SummarizedOffenseDescription` for the column and `is in` for the opreator. On the Values, pick `CAR PROWL`, `BURGLARY`, `PROPERTY DAMAGE`, `WARRANT ARREST`, and `STOLEN PROPERTY`. Go back to `Chart` and examine the boxplot with only the top five crime types.  

<br>


### 5. Third part - Exploring multiple variables (only numeric)

Now that we explored the two variables, we can move onto exploring multiple variables. For this, choose `Analytics`, and select `Correlation by Columns`. For the colums, select the following columns:<br>

<ul>
<li>total_crime</li>
<li>pop</li>
<li>pop_den</li>
<li>p_kids</li>
<li>p_seniors</li>
<li>p_female</li>
<li>p_white</li>
<li>some_college</li>
<li>med_hh_inc</li>
<li>p_poverty</li>
</ul>

When you are done selecting the variables, click Run. We can see a correlation plot.


<br><br>


## Use the EDA techniques to explore your city's data

1. Create a bar plot of crime types and total crime. Sort the plot by y axis. <br>

2. Create a scatter plot of pop and total_crime. <br>

3. Create a correlation plot of at least 5 key variables of your choice <br>

4. Create a short report that includes the above three plots and send the document to the course email (urbanbigdata2019@gmail.com). 
 + <font color="red"><b>[IMPORTANT]</b></font> Please use the following email title format: <br>
 `VSP BigData [lecture number] - [group number] - [presenter name]` <br>
 ex), `VSP BigData Lecture 7 - Group 1 - Bill Gates` <br>
 
<br><br><br>
 

 




